---
title: "An Introduction to dplyr"
author: "Phil Canham"
date: "30th October 2018"
output: word_document
---

# An Introduction to dplyr

Dplyr is one of the most useful R librarys for manipulating data. Think of it like a Swiss Army knife for data.

## What does it do? 

Once we have a data set to work with, almost always it is not in the format, or shape that we want it to be. In Excel we might for example use filters, formulas, lookups and pivot tables to make the data more useful. In R we can use the tools in the dplyr libarary to do these things and much more.

The other important thing to remember is that in R we normally provide a series of instructions which can then always be repeated and changed without having to start again each time, so we have total control of what happens to the data, and the original data should never be permanently overwritten unless we explicity ask R to do this (which is generally bad practice)

The author of the library, Hadley Wickham,  has streamlined the process of data manipulation by following a common format using the concept of verbs, (ie words used to describe actions). The key dplyr verbs we will consider today are:

* filter
* select
* arrange
* mutate
* group_by
* summarise

### Lets try them out

### First we clear the memory and any previously loaded work from R Studio

```{r clear space}
rm(list= ls())
```


```{r setup, echo=F}
knitr::opts_knit$set(root.dir = '//dsfin/corpcom/LBBDR/TNG/Session 3') ### this sets the directory for the whole markdown document
```

### Second, set the working directory to allow us to load in our data

```{r}
setwd('//dsfin/corpcom/LBBDR/TNG/Session 3')
```

### We can now load any librarys we are interested in using - here we load the dplyr library


```{r loadlibrary, warning=FALSE, echo=TRUE, results=TRUE}

library(dplyr)
# I also want to use another library which prints out much neater tables but we dont need to do this
library(tibble)
```

### The next step is to read in the data from our csv file

```{r load data}
flights <- read.csv('session3a.csv')
```
### and have a look at a summary of the data we have loaded

```{r summaryStats}
summary(flights)
```

This dataset contains all flights departing from Houston airports IAH (George Bush Intercontinental)
and HOU (Houston Hobby) in 2011. The data comes from the Research and Innovation Technology
Administration at the Bureau of Transportation statistics: http://www.transtats.bts.gov/DatabaseInfo.asp?DB_ID=120&Link=0

Information on the variables and the dataset can be found [here](https://cran.r-project.org/web/packages/hflights/hflights.pdf "Symbol Types")

There is also a pdf of this document in the working directory folder




### now check the top few rows of the data

```{r head}
head(flights)
```


### lets use the tibble library to convert our data frame into a format that prints out nicely

```{r tibble}
flights <- as.tibble(flights) ## remember we dont need to do this step it is just for better display
```


## filter: Keep rows matching criteria

* Base R approach to filtering forces you to repeat the data frame's name
* dplyr approach is simpler to write and read
* Command structure (for all dplyr verbs):
    * first argument is a data frame
    * return value is a data frame
    * nothing is modified in place
* Note: dplyr generally does not preserve row names

```{r results='hide'}
# base R approach to view all flights on January 1
flights[flights$Month==1 & flights$DayofMonth==1, ]
```

```{r}
# dplyr approach
# note: you can use comma or ampersand to represent AND condition
filter(flights, Month==1, DayofMonth==1)

# use pipe for OR condition
filter(flights, UniqueCarrier=="AA" | UniqueCarrier=="UA")
```

```{r results='hide'}
# you can also use %in% operator
filter(flights, UniqueCarrier %in% c("AA", "UA"))
```


## select: Pick columns by name

* Base R approach is awkward to type and to read
* dplyr approach uses similar syntax to `filter`
* Like a SELECT in SQL

```{r results='hide'}
# base R approach to select DepTime, ArrTime, and FlightNum columns
flights[, c("DepTime", "ArrTime", "FlightNum")]
```

```{r}
# dplyr approach
select(flights, DepTime, ArrTime, FlightNum)

# use colon to select multiple contiguous columns, and use `contains` to match columns by name
# note: `starts_with`, `ends_with`, and `matches` (for regular expressions) can also be used to match columns by name
select(flights, Year:DayofMonth, contains("Taxi"), contains("Delay"))
```


## "Chaining" or "Piping"

* One way to perform multiple operations in one line is by nesting but this can be quite complex to follow
* We can also write commands in a natural order by using the `%>%` pipe operator (which can be pronounced as "then")

for example

eat food %>% drink beer

eat food THEN drink beer


```{r results='hide'}
# nesting method to select UniqueCarrier and DepDelay columns and filter for delays over 60 minutes
filter(select(flights, UniqueCarrier, DepDelay), DepDelay > 600)
```

```{r}
# chaining or piping method
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    filter(DepDelay > 600)
```

* Chaining increases readability significantly when there are many commands
* Operator is automatically imported from the [magrittr](https://github.com/smbache/magrittr) package
* %>% Can be used to replace nesting in R commands outside of dplyr
* the shortcut key is ctrl/shift M

```{r results='hide'}
# create two vectors and calculate Euclidian distance between them. This is the less readable way without chaining
x1 <- 1:5; x2 <- 2:6
sqrt(sum((x1-x2)^2))
```

```{r}
# chaining method
(x1-x2)^2 %>% sum() %>% sqrt()
```





## arrange: Reorder rows

```{r results='hide'}
# base R approach to select UniqueCarrier and DepDelay columns and sort by DepDelay
flights[order(flights$DepDelay), c("UniqueCarrier", "DepDelay")]
```

```{r}
# dplyr approach
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(DepDelay)
```

```{r results='hide'}
# use `desc` for descending
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay))
```



## mutate: Add new variables

* Create new variables that are functions of existing variables

```{r results='hide'}
# base R approach to create a new variable Speed (in mph)
flights$Speed <- flights$Distance / flights$AirTime*60
flights[, c("Distance", "AirTime", "Speed")]
```

```{r}
# dplyr approach (prints the new variable but does not store it)
flights %>%
    select(Distance, AirTime) %>%
    mutate(Speed = Distance/AirTime*60)

# store the new variable
flights <- flights %>% mutate(Speed = Distance/AirTime*60)

# look at the names of the variables in flights

names(flights) # we can see that "Speed" has no been added to the dataframe at the end

```



## summarise: Reduce variables to values

* Primarily useful with data that has been grouped by one or more variables
* Think of pivot tables in Excel as a comparable idea
* `group_by` creates the groups that will be operated on
* `summarise` uses the provided aggregation function to summarise each group

```{r results='hide'}
# base R approaches to calculate the average arrival delay to each destination
head(with(flights, tapply(ArrDelay, Dest, mean, na.rm=TRUE)))
head(aggregate(ArrDelay ~ Dest, flights, mean))
```

```{r}
# dplyr approach: create a table grouped by Dest, and then summarise each group by taking the mean of ArrDelay
flights %>%
    group_by(Dest) %>%
    summarise(avg_delay = mean(ArrDelay, na.rm=TRUE)) ### the na.rm = TRUE tells R to ignore any missing values  to avoid an error
```

### There are three other variants of `summarise`
* _all affects every variable
* _at affects variables selected with a character vector or vars()
* _if affects variables selected with a predicate function:

### So for example:
* `summarise_at` allows you to apply the same summary function to multiple columns at once
* Note: `mutate_at` is also available

```{r}
# for each carrier, calculate the percentage of flights cancelled or diverted
flights %>%
    group_by(UniqueCarrier) %>%
    summarise_at(vars(Cancelled, Diverted), funs(mean))

#### so the above can be read as "  for the dataframe called flights, group by the Unique Carrier id THEN summarise these groups
# using the variables Cancelled and Diverted using the MEAN function "

# we can also do more than one summary

# for each carrier, calculate the minimum and maximum arrival and departure delays
flights %>%
    group_by(UniqueCarrier) %>%
    summarise_at(vars(ArrDelay, DepDelay), funs(min, max), na.rm = TRUE)
  
```


* Helper function `n()` counts the number of rows in a group
* Helper function `n_distinct(vector)` counts the number of unique items in that vector

```{r}
# for each day of the year, count the total number of flights and sort in descending order
flights %>%
    group_by(Month, DayofMonth) %>%
    summarise(flight_count = n()) %>%
    arrange(desc(flight_count))

# rewrite more simply with the `tally` function
flights %>%
    group_by(Month, DayofMonth) %>%
    tally(sort = TRUE)

# for each destination, count the total number of flights and the number of distinct planes that flew there
flights %>%
    group_by(Dest) %>%
    summarise(flight_count = n(), plane_count = n_distinct(TailNum))
```

* Grouping can sometimes be useful without summarising

```{r}
# for each destination, show the number of cancelled and not cancelled flights
flights %>%
    group_by(Dest) %>%
    select(Dest,Cancelled) %>%
    table() %>%  #### table() is actually a base R function
    head(20) ### show top 20 rows
```

